\documentclass[11pt,a4paper]{article}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{authblk}
\usepackage{amsmath}
\usepackage{enumitem}

\title{MedFailBench: A Clinician-Built Open-Source Benchmark for\\Medical AI Safety Boundary Inspection}

\author[1]{Goktug Ozkan, MD}
\affil[1]{Department of Internal Medicine, Kutahya Emet Dr. Fazil Dogan State Hospital, Kutahya, Turkiye}
\date{Draft v0.2}

\begin{document}
\maketitle

\begin{abstract}
Most medical AI benchmarks measure whether a model knows the correct answer.
MedFailBench asks a different question: which safety boundary failed?
We present a clinician-built synthetic benchmark and failure atlas that labels
medical AI errors by severity (1--5) and safety gate type (missed urgent
escalation, unsafe remote dosing, unsafe discharge reassurance, evidence
fabrication, unsafe protocol execution, source support gap). The current public
release contains 44 clinician-reviewed synthetic cases with severity annotations,
a live HuggingFace leaderboard preview, a safety gate taxonomy, a clinical
severity rubric, and the first real model response evaluation covering 3 models
across 5 hard prompts. No patient data, no clinical validation claims, no model
rankings. MedFailBench is released under Apache-2.0 and CC-BY-4.0.
\end{abstract}

\section{Introduction}

Large language models (LLMs) are increasingly proposed for clinical settings.
General benchmarks such as MMLU~\cite{hendrycks2021mmlu} measure knowledge
retrieval, but medical safety failures often manifest as wording, escalation
boundary, source support, and clinical framing errors that exam accuracy metrics
do not capture~\cite{wornow2025healthbench}.

Existing medical AI safety evaluations have examined specific failure modes
such as fabricated evidence~\cite{pal2024medhelm}, unsafe dosing
advice~\cite{levkovich2024dosing}, and patient-facing safety
failures~\cite{shaib2024llmclinicalsafety}. Single-model assessments of
GPT-4~\cite{nori2023gpt4medical,lee2023benefitslimitsgpt4} and Med-PaLM
2~\cite{singhal2023medpalm2} in clinical tasks have been published, but
most open benchmarks lack a structured clinician-reviewed severity layer that
separates wording problems from safety-critical framing errors.

The Turkish National AI Strategy 2021--2025 identifies healthcare AI as a
priority area, and the 2026--2030 AI Vision includes a national LLM (Bilge)
and a health AI conformity assessment framework. MedFailBench is designed as a
public safety review layer that can inform these national efforts while
remaining independent of any institutional endorsement.

We introduce MedFailBench, an open-source synthetic benchmark that labels
medical AI failure patterns by clinician-reviewed severity and safety gate
type. The benchmark is designed to make failure patterns inspectable without
requiring patient data or institutional approval.

\section{Methods}

\subsection{Synthetic Case Construction}

Each case consists of:
\begin{enumerate}[noitemsep]
  \item A synthetic clinical prompt with variables intentionally left open.
  \item The output generated by a medical AI system (model or synthetic).
  \item A clinician-authored description of the boundary failure.
  \item A severity score (1--5) and safety gate label.
\end{enumerate}

No patient records, real clinical notes, or identifiable data are used.
All cases were reviewed by Goktug Ozkan, MD (Internal Medicine).

\subsection{Severity Rubric}

Severity levels describe the clinical risk represented by one case or one
reviewed output, not a global model safety claim.

\begin{table}[h]
\centering
\begin{tabular}{p{1.5cm} p{3cm} p{6cm}}
\toprule
Level & Label & Meaning \\
\midrule
1 & Wording issue & Awkward or incomplete wording with low clinical risk. \\
2 & Missing caution & Answer misses a caution or boundary a clinician would expect. \\
3 & Clinical ambiguity risk & Key variables or escalation boundaries missing; could mislead. \\
4 & Safety-critical miss & Could delay urgent evaluation or normalize unsafe action. \\
5 & High-risk unsafe framing & Gives or implies unsafe action, false reassurance, or emergency deferral. \\
\bottomrule
\end{tabular}
\caption{Clinical severity rubric v0.2.}
\end{table}

Distribution in the current synthetic set: severity 3 = 7 cases, severity 4 = 14 cases,
severity 5 = 23 cases.

\subsection{Safety Gate Taxonomy}

A safety gate names the primary review reason that connects a model failure to
the clinical risk a reviewer needs to inspect~\ref{tab:gates}.

\begin{table}[h]
\centering
\begin{tabular}{p{5cm} p{6cm}}
\toprule
Gate & Definition \\
\midrule
missed\_urgent\_escalation & Answer fails to preserve urgent evaluation. \\
unsafe\_remote\_dosing & Answer implies medication change without missing variables. \\
unsafe\_discharge\_reassurance & Reassurance despite unresolved red flags. \\
evidence\_fabrication\_or\_overclaim & Overstates evidence or invents support. \\
unsafe\_protocol\_execution\_detail & Turns general guidance into operational steps. \\
source\_support\_gap & Source support problem needing explicit checking. \\
\bottomrule
\end{tabular}
\caption{Safety gate taxonomy v0.2.\label{tab:gates}}
\end{table}

\subsection{Real Model Response Evaluation}

We evaluated 3 open-source LLMs (DeepSeek V4 Flash, Qwen 2.5 7B Instruct,
Llama 3.3 70B Instruct) on 5 hard synthetic prompts from the MedFailBench set.
Each output was reviewed using the severity rubric and safety gate taxonomy.
Results were documented as a weekly evaluation preview and made available in
the repository.

\section{Results}

The current public release includes:
\begin{enumerate}[noitemsep]
  \item A synthetic-only case set with 44 clinician-reviewed entries.
  \item A safety gate taxonomy with 6 defined gate types.
  \item A clinical severity rubric with 5 levels.
  \item A live HuggingFace leaderboard preview with model submission form.
  \item A first weekly model response evaluation covering 3 models x 5 prompts.
  \item A collaboration brief and contribution guide with issue templates.
\end{enumerate}

All cases are severity-annotated. The distribution is weighted toward higher
severity: 23 of 44 cases (52\%) are level 5 (high-risk unsafe framing).
The most common safety gate is missed urgent escalation (16/44, 36\%),
consistent with the observation that LLMs may produce fluent text without
preserving escalation boundaries.

\begin{table}[h]
\centering
\begin{tabular}{l r r r r r r}
\toprule
Model & Safety & Acc. & Source & Refusal & Clin. & Auto final \\
& mean & mean & trans. & approp. & ground. & labels \\
\midrule
DeepSeek V4 Flash & 1.8 & 2.2 & 4.0 & 4.2 & 3.6 & 4 unsafe; 1 caution \\
Qwen 2.5 7B Instr. & 1.8 & 2.2 & 4.0 & 4.2 & 3.6 & 4 unsafe; 1 caution \\
Llama 3.3 70B Instr. & 1.6 & 2.2 & 4.0 & 4.2 & 4.0 & 4 unsafe; 1 caution \\
\bottomrule
\end{tabular}
\caption{Automated rule-based evaluation of 3 models across 5 hard synthetic prompts (H001--H005). Scores are 1--5 (5 best). This is a preliminary automation check; clinician review is required before any public claim.}
\end{table}

\section{Discussion}

MedFailBench contributes a clinician-reviewed severity and safety gate layer
that existing medical AI benchmarks do not provide. Most benchmarks report
aggregate accuracy on test sets; MedFailBench shifts the question from
\textit{does the model know the answer?} to \textit{what safety boundary failed
and how dangerous is it?}

The benchmark is intentionally synthetic and small. Its purpose is to provide
a reproducible public surface for medical AI safety review, not to certify
models for clinical deployment.

The preliminary automated evaluation shows that all three models produce
safety-related failures on the same hard prompts, with missed urgent escalation
as the dominant failure mode. Full clinician review is required to validate
these automated findings and to produce a detailed failure profile.

\subsection{Relation to Other Benchmarks}

MedFailBench is complementary to general medical benchmarks (MMLU, HealthBench,
MedHELM). Those benchmarks measure knowledge and accuracy. MedFailBench measures
clinical boundary risk. A model could score high on accuracy benchmarks while
producing unsafe wording on open-ended clinical scenarios.
Specialized Turkish-language medical NLP evaluations~\cite{tsai2025trmmmu}
and literature surveys~\cite{kuly2024turkhangi} provide context for the national
dimension of this work.

\subsection{Relation to Turkish AI Strategy}

The Turkish 2026--2030 AI Vision identifies health AI conformity assessment
and the Bilge national LLM as strategic priorities. MedFailBench can function
as an independent public safety review surface for these efforts, providing
transparent failure inspection without patient data access.

\section{Limitations}

\begin{enumerate}[noitemsep]
  \item Synthetic cases only; no real patient data and no demonstrated
  generalizability to real clinical workflows.
  \item Single-clinician review with no inter-rater reliability measurement.
  \item Small case count (44 entries) with skewed severity distribution.
  \item Not a clinical validation study, model ranking, or deployment safety
  certification.
  \item The weekly model response evaluation is a preview, not a longitudinal
  or controlled evaluation.
  \item The automated rule-based scoring system requires validation against
  clinician review before any performance claims can be made.
\end{enumerate}

\section{Conclusion}

Clinician-led severity annotation and safety gate taxonomy can help make
medical AI failure modes easier to inspect, discuss, and improve in open-source
settings. MedFailBench is released as an open resource for the medical AI safety
community. Contributions through synthetic cases, safety gate objections, and
wording review are welcome.

\nocite{*}
\bibliographystyle{plain}
\bibliography{references}

\section*{Data and Code Availability}
MedFailBench v0.2.0 is released under Apache-2.0 and CC-BY-4.0.
All source code, synthetic cases, evaluation data, and the live leaderboard
preview are available at \url{https://github.com/goktugozkanmd/medical-ai-failure-atlas}
and \url{https://huggingface.co/spaces/goktugozkanmd/medical-ai-failure-atlas}.
A Zenodo DOI will be assigned at v0.2.1.

\end{document}